Introduction

### 2024-04-12 Sections/Introduction.tex
adding more citations to intro

% ---------------------------
% END CUSTOM PREAMBLE STUFF
% ---------------------------
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{sidenotes}{}{\usepackage{sidenotes}}
\@ifpackageloaded{marginnote}{}{\usepackage{marginnote}}
\makeatother
\makeatletter
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Full Dissertation},
  colorlinks=true,
  linkcolor={DarkSlateBlue},
  filecolor={Maroon},
  citecolor={DarkSlateBlue},
  urlcolor={DarkSlateBlue},
  pdfcreator={LaTeX via pandoc}}

@@

% SHORT TITLE           Page #
\fancypagestyle{normal}{
  \fancyhf{}
  \lhead{\uppercase{ Full Dissertation }}
  \rhead{\thepage}
}

% Running head: SHORT TITLE        Page #
\fancypagestyle{title}{
  \fancyhf{}
  \lhead{Running head: \uppercase{ Full Dissertation }}
  \rhead{}
}

@@

\title{Full Dissertation:}

\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par #1 \par}{}{}
}
\makeatother
\subtitle{An instance-based model account of the benefits of varied
practice in visuomotor skill}

@@

\begin{center}
\singlespacing
\textbf{Full Dissertation:\\An instance-based model account of the
benefits of varied practice in visuomotor skill}
\end{center}

@@

\renewcommand*\contentsname{Contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}
\setstretch{2}
\section{Variability and
Generalization}\label{variability-and-generalization}

The factors that influence the generalization of learning are of
considerable interest to both researchers exploring the human learning
system and practitioners aiming to enhance the effectiveness of
educational and training interventions. The present effort will focus
specifically on the role of variability in the learning input.
Variability manipulations typically regulate either the number of
distinct instances presented to learners during training, or the
dispersion of these instances. Such manipulations have been empirically
demonstrated to affect subsequent generalization performance. This essay
will offer an in-depth review of the extant literature on the influence
of variability, spanning multiple relevant domains.

@@

Studies investigating the ``benefits of variability'' hypothesis usually
assign participants to either a constant or varied group for the
training stage of the experiment. Then, subjects in both groups complete
an identical testing stage which often consists items/conditions seen
during training, and novel items/conditions. If the varied group
performs better in the testing stage, this is taken for evidence of the
benefits of variability hypothesis. Even within this relatively
straightforward between-groups design, researchers must navigate several
crucial methodological choices, highlighted below:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  Variables Subject to Variation. In multidimensional tasks, researchers
  have the option to vary numerous variables. The experimenters must
  decide the specific dimension(s) across which variation will occur.
  For instance, in a projectile throwing accuracy task -- researchers
  might vary the distance from the target, the size of the target, the
  weight of the projectile. They might also vary a contextual variable
  not directly relevant to the task, but which will still be encoded by
  the subject on a trial by trial basis, e.g.~the background color.
\item
  Magnitude of Variation, relative to the control condition. The
  simplest comparison would be to compare a constant group who trains
  with 1 example/condition, against a varied group that trains from 2
  examples/conditions. However, it is not uncommon in the literature for
  the varied condition to train from 3 or 4 conditions. For example,
  \textcite{catalanoDistantTransferCoincident1984a} train varied
  subjects from 4 different velocities in their coincident timing task,
  and \autocite{goodeSuperiorityVariableRepeated2008} have varied
  subjects' practice with 3 different variants (i.e.~different letter
  scrambles of the same word) of an anagram for a given word, while
  their constant participants view the same variant 3 different times.
  Alternatively, rather than a constant vs.~varied comparison, subjects
  in all conditions might experience a variety of training items, but
  with one group experiencing a greater number of unique items
  \autocite{nosofskyModelguidedSearchOptimal2018}.
\item
  Locations within Task-Space. For tasks in which the stimuli or
  conditions fall within a continuous metric space, the experimenter
  must decide whether the varied instances are relatively close together
  (e.g.~throwing a ball from a distance of 4 feet and 5 feet), or far
  apart (throwing from 4 feet and 20 feet). Spreading the varied
  training items further apart may be beneficial in terms of providing a
  more representative sample of the task space to the learner, however
  large distances may also result in significant differences in
  difficulty between the training examples, which can be a common
  confound in variability studies.
\item
  Proximity of Testing to Training Conditions. Intuitively, the fairest
  form of comparison is to include testing conditions that are of an
  equivalent distance from both the varied and constant groups. However
  researchers might also attempt to demonstrate the benefits of
  variation as being sufficiently powerful to outperform constant
  training, even in cases where the constant group trained from a closer
  proximity to the testing conditions, or whose training conditions are
  identical to the testing conditions
  \autocite{goodeSuperiorityVariableRepeated2008,kerrSpecificVariedPractice1978}.
\end{enumerate}

\subsection{Variability Literature
Review}\label{variability-literature-review}

An early and influential work on the influence of variability on
category learning is that of \textcite{posnerGenesisAbstractIdeas1968}.
In an ambitious attempt to address the question of how category
information is represented, the authors trained participants to
categorize artificial dot patterns, manipulating whether learners were
exposed to examples clustered close to the category prototypes (e.g.~a
low variability condition), or spread further away from the prototype
(the varied-training group). It should be noted that both groups in this
study were trained with the same number of unique instances and the
manipulated difference was how spread out the instances were. The
authors claim based on prior experiments using the same stimuli, that
the training stimuli for the varied group were at least as far away from
the testing stimuli as the training stimuli of the less-varied group.
The authors interpreted their findings as evidence for the extraction of
an abstraction or schema that is extracted and stored, and then over
time becomes more likely to be the reference point from which
generalization occurs, given that specific instances are thought to
decay at a faster rate than prototypes or schema. The Posner and Keele
study has been extremely influential and continues to be cited in
contemporary research as clear evidence that schema abstraction
underlies the benefits of varied training. It's also referenced as a key
influence in the development of ``Schema Theory of Motor Learning''
\autocite{schmidtSchemaTheoryDiscrete1975}, which in turn influenced
decades of investigations on the potential benefits of varied training
in motor skill learning. However, the classic Posner \& Keele study
despite being far more carefully designed than many subsequent studies,
and despite being a relative rarity in explicitly discussing and
attempting to control for potential confounds of similarity between
groups, may nevertheless be emblematic of a common issue in many
investigations of the effects of varied training on learning. The
problem with Posner \& Keele's conclusion was demonstrated clearly
almost 3 decades later \autocite{palmeriCentralTendenciesExtreme2001},
when researchers conducting a near replication of the original study
also collected similarity judgements following training and performed
multidimensional scaling analysis. Rather than being in the middle of
the training stimuli as was the case in the physical stimuli space, the
psychological representation of the prototype was shown reside at an
extreme point, and generalization patterns by participants that would
have seemed to warrant the learning of a prototype were then easily
accounted for with only the assumption that the participants encoded
instances. One of the primary concerns of the present paper is that many
of the studies which purport to explain the benefits of variation via
prototypes, schemas, or other abstractions, are often overlooking the
potential of instance based similarity accounts.

\subsection{Motor skill learning}\label{motor-skill-learning}

Training variation has also been shown to promote transfer in motor
learning. Much of this research has been influenced by the work of
\autocite{schmidtSchemaTheoryDiscrete1975}, who proposed a schema-based
account of motor learning as an attempt to address the longstanding
problem of how novel movements are produced. Schema theory presumes a
priori that learners possess general motor programs for classes of
movements, such as an underhand throw. When called up for use, such
programs must be parameterized, as well as schema rules that determine
how a motor program is parameterized or scaled for a particular
movement. Schema theory predicts that varied training results in the
formation of a more general schema-rule, which can allow for transfer to
novel movements within a given movement class, such as an underhand
throw (though it is agnostic to the development of the movement classes
themselves). Experiments that test this hypothesis are often designed to
compare the transfer performance of a constant-trained group against
that of a varied-trained group. Both groups train on the same task, but
the varied group practices with multiple instances along some
task-relevant dimension that remains invariant for the constant group.
For example, investigators might train two groups of participants to
throw a projectile at a target, with a constant group that throws from a
single location, and a varied group that throws from multiple locations.
Both groups are then tested from novel locations.

One of the earliest, and still often cited investigations of Schmidt's
benefits of variability hypothesis was the work of
\textcite{kerrSpecificVariedPractice1978}. Two groups of children, aged
8 and 12, were assigned to either constant or varied training of a bean
bag throwing task. The constant group practiced throwing a bean-bag at a
small target placed 3 feet in front of them, and the varied group
practiced throwing from a distance of both 2 feet and 4 feet.
Participants were blindfolded and unable to see the target while making
each throw but would receive feedback by looking at where the beanbag
had landed in between each training trial. 12 weeks later, all of the
children were given a final test from a distance of 3 feet which was
novel for the varied participants and repeated for the constant
participants. Participants were also blindfolded for testing and did not
receive trial by trial feedback in this stage. However, at the halfway
point of the testing stage they were allowed to see the landing location
of the 4 beanbags they had thrown, and then completed the final 4
testing throws. In both age groups, participants performed significantly
better in the varied condition than the constant condition, though the
effect was larger for the younger, 8-year-old children. Although this
design does not directly assess the hypothesis of varied training
producing superior generalization to constant training (since the
constant group is not tested from a novel position), it nevertheless
offers a compelling example of the merits of varied practice.

On occasion the Kerr and Booth design may be nested within a larger
experimental design. One such study that used a movement timing task,
wherein subjects had to move their hand from a starting location, to a
target location, attempting to arrive at the target location at specific
time following the onset of a cue
\autocite{wrisbergVariabilityPracticeHypothesis1987}. This study
utilized 4 different constant groups, and 3 varied groups, with one of
the constant groups training under conditions identical to the testing
conditions, and which were not trained on by any of the varied groups,
e.g.~the design of Kerr and Booth. However, in this case the varied
group did not outperform the constant group. A more recent study
attempting a slightly more direct replication of the original Kerr \&
Booth study \autocite{willeyLongtermMotorLearning2018}, having subjects
throw beanbags at a target, with the varied group training from
positions (5 and 9 feet) on either side of the constant group (7 feet).
However, this study diverged from the original in that the participants
were adults; they faced away from the target and threw the beanbag
backwards over their bodies; they alternated using their right and left
hands every 6 trials; and underwent a relatively extreme amount of
training (20 sessions with 60 practice trials each, spread out over 5-7
weeks). Like \textcite{wrisbergVariabilityPracticeHypothesis1987}, this
study did not find a varied advantage from the constant training
position, though the varied group did perform better at distances novel
to both groups.

Some support for the Kerr and Booth findings was found with a relatively
less common experimental task of training participants in hitting a
projectile at a target with the use of a racket
\autocite{greenPracticeVariabilityTransfer1995a}. Varied participants
trained with tennis, squash, badminton, and short-tennis rackets were
compared against constant subjects trained with only a tennis racket.
One of the testing conditions had subjects repeat the use of the tennis
racket, which had been used on all 128 training trials for the constant
group, and only 32 training trials for the varied group. Nevertheless,
the varied group outperformed the constant group when using the tennis
racket at testing, and also performed better in conditions with several
novel racket lengths. Of course, this finding is less surprising than
that of Kerr \& Booth, given that varied subjects did have some prior
exposure to the constant groups condition. This highlights an issue
rarely discussed in the literature, of how much practice from an
additional position might be necessary to induce benefits. Experimenters
almost uniformly have varied participants train with an equivalent
number of trials from each of their conditions

One of the few studies that has replicated the surprising result of
varied outperforming constant, from the constant training condition, did
so in the relatively distant domain of verbal manipulation
\autocite{goodeSuperiorityVariableRepeated2008}. All participants
trained to solve anagrams of 40 different words ranging in length from 5
to 11 letters, with an anagram of each word repeated 3 times throughout
training, for a total of 120 training trials. Although subjects in all
conditions were exposed to the same 40 unique words (i.e.~the solution
to an anagram), participants in the varied group saw 3 different
arrangements for each solution-word, such as DOLOF, FOLOD, and OOFLD for
the solution word FLOOD, whereas constant subjects would train on three
repetitions of LDOOF (spread evenly across training). Two different
constant groups were used. Both constant groups trained with three
repetitions of the same word scramble, but for constant group A, the
testing phase consisted of the identical letter arrangement to that seen
during training (e.g.~LDOOF), whereas for constant group B, the testing
phase consisted of a arrangement they had not seen during training, thus
presenting them with a testing situation similar situation to the varied
group. At the testing stage, the varied group outperformed both constant
groups, a particularly impressive result, given that constant group A
had 3 prior exposures to the word arrangement (i.e.~the particular
permutation of letters) which the varied group had not explicitly seen.
However varied subjects in this study did not exhibit the typical
decrement in the training phase typical of other varied manipulations in
the literature, and actually achieved higher levels of anagram solving
accuracy by the end of training than either of the constant groups --
solving 2 more anagrams on average than the constant group. This might
suggest that for tasks of this nature where the learner can simply get
stuck with a particular word scramble, repeated exposure to the
identical scramble might be less helpful towards finding the solution
than being given a different arrangement of the same letters. This
contention is supported by the fact that constant group A, who was
tested on the identical arrangement as they experienced during training,
performed no better at testing than did constant group B, who had
trained on a different arrangement of the same word solution -- further
suggesting that there may not have been a strong identity advantage in
this task.

Pitting varied against constant practice against each other on the home
turf of the constant group provides a compelling argument for the
benefits of varied training, as well as an interesting challenge for
theoretical accounts that posit generalization to occur as some function
of distance. However, despite its appeal this particular contrast is
relatively uncommon in the literature. It is unclear whether this may be
cause for concern over publication bias, or just researchers feeling the
design is too risky. A far more common design is to have separate
constant groups that each train exclusively from each of the conditions
that the varied group encounters
\autocite{catalanoDistantTransferCoincident1984a,chuaPracticeVariabilityPromotes2019,newellVariabilityPracticeTransfer1976,moxleySchemaVariabilityPractice1979,mccrackenTestSchemaTheory1977},
or for a single constant group to train from just one of the conditions
experienced by the varied participants
\autocite{pigottMotorSchemaStructure1984,rollerVariablePracticeLenses2001,wrisbergTrainingProductionNovel1984,wrisbergDevelopingCoincidentTiming1983}.
A less common contrast places the constant group training in a region of
the task space outside of the range of examples experienced by the
varied group, but distinct from the transfer condition
\autocite{wrisbergVariabilityPracticeHypothesis1987,wulfVariabilityPracticeImplicit1997}.

Of particular relevant to the current essay is the early work of
\textcite{catalanoDistantTransferCoincident1984a}, as theirs was one of
the earliest studies to investigate the influence of varied vs.~constant
training on multiple testing locations of graded distance from the
training condition. Participants were trained on coincident timing task,
in which subjects observe a series of lightbulbs turning on sequentially
at a consistent rate and attempt to time a button response with the
onset of the final bulb. The constant groups trained with a single
velocity of either 5,7,9, or 11 mph, while the varied group trained from
all 4 of these velocities. Participants were then assigned to one of
four possible generalization conditions, all of which fell outside of
the range of the varied training conditions -- 1, 3, 13 or 15 mph. As is
often the case, the varied group performed worse during the training
phase. In the testing phase, the general pattern was for all
participants to perform worse as the testing conditions became further
away from the training conditions, but since the drop off in performance
as a function of distance was far less steep for the varied group, the
authors suggested that varied training induced a decremented
generalization gradient, such that the varied participants were less
affected by the change between training and testing conditions.

\subsection{Category learning}\label{category-learning}

>>>>

% ---------------------------
% END CUSTOM PREAMBLE STUFF
% ---------------------------
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Introduction},
  colorlinks=true,
  linkcolor={DarkSlateBlue},
  filecolor={Maroon},
  citecolor={DarkSlateBlue},
  urlcolor={DarkSlateBlue},
  pdfcreator={LaTeX via pandoc}}

@@

% SHORT TITLE           Page #
\fancypagestyle{normal}{
  \fancyhf{}
  \lhead{\uppercase{ Introduction }}
  \rhead{\thepage}
}

% Running head: SHORT TITLE        Page #
\fancypagestyle{title}{
  \fancyhf{}
  \lhead{Running head: \uppercase{ Introduction }}
  \rhead{}
}

@@

\title{Introduction}

@@

\begin{center}
\singlespacing
\textbf{Introduction}
\end{center}

@@

\setstretch{2}
\section{Introduction}\label{introduction}

The past century of research on human learning has produced ample
evidence that although learners can improve at almost any task, such
improvements are often specific to the trained task, with unreliable or
even nonexistent transfer or generalization to novel tasks or conditions
\autocite{barnettWhenWhereWe2002,dettermanCaseProsecutionTransfer1993}.
Such generalization challenges are of noteworthy practical relevance,
given that educators, trainers, and rehabilitators typically intend for
their students to be able to apply what they have learned to new
situations. It is therefore important to better understand the factors
that influence generalization, and to develop cognitive models that can
predict when generalization is likely to occur. Such characteristics
have included training difficulty, spacing, temporal order, feedback
schedules, and the primary focus of the current work - the variability
of training examples.

@@

Varied training has been shown to influence learning in wide array of
different tasks and domains, including categorization
\autocite{hahnEffectsCategoryDiversity2005,maddoxStimulusRangeDiscontinuity2011,posnerGenesisAbstractIdeas1968,nosofskyModelguidedSearchOptimal2019,morgensternOneshotCategorizationNovel2019,plebanekEffectsFrequencyVariability2021},
language learning
\autocite{jonesDensityDistinctivenessEarly2020,perryLearnLocallyThink2010,twomeyAllRightNoises2018,wonnacottInputEffectsAcquisition2012,brekelmansDoesHighVariability2022},
anagram completion \autocite{goodeSuperiorityVariableRepeated2008},
trajectory extrapolation
\autocite{fulvioTaskSpecificResponseStrategy2014}, cognitive control
tasks
\autocite{sabahWhenLessMore2019,moshon-cohenStimulusVariabilityImproves2024},
associative learning
\autocite{leeEvidentialDiversityIncreases2019,reichmannVariabilityAbstractionEvaluative2023,fanStimulusDiversityIncreases2022,pradaExperiencedCategoryVariability2020},
visual search
\autocite{georgeStimulusVariabilityTask2021,gonzalezDiversityTrainingEnhances2011,kelleyLearningAttendEffects2009},
voice identity learning \autocite{lavanEffectsHighVariability2019}, face
recognition
\autocite{honigPerceptualSimilarityModulates2022,burtonIdentityVariationRepresentations2016,menonVariationPhotosSame2015},
the perception of social group heterogeneity
\autocite{linvilleExemplarAbstractionModels1993,konovalovaInformationSamplingExplanation2020,parkPerceptionVariabilityCategory1987,gershmanStructureLearningPrinciples2023}
, simple motor learning
\autocite{braunMotorTaskVariation2009,kerrSpecificVariedPractice1978,rollerVariablePracticeLenses2001,willeyLimitedGeneralizationVaried2018},
sports training
\autocite{greenPracticeVariabilityTransfer1995a,northEffectConsistentVaried2019,breslinConstantVariablePractice2012},
and complex skill learning
\autocite{seowTransferEffectsVaried2019,huetEducationAttentionExplanation2011,hacquesVisualControlClimbing2022}.
See \textcite{czyzVariabilityPracticeInformation2021} or
\textcite{ravivHowVariabilityShapes2022} for more detailed reviews.

Research on the effects of varied training typically manipulates
variability in one of two ways. In the first approach, a high
variability group is exposed to a greater number of unique instances
during training, while a low variability group receives fewer unique
instances with more repetitions. Alternatively, both groups may receive
the same number of unique instances, but the high variability group's
instances are more widely distributed or spread out in the relevant
psychological space, while the low variability group's instances are
clustered more tightly together. Researchers then compare the training
groups in terms of their performance during the training phase, as well
as their generalization performance during a testing phase. Researchers
will usually compare the performance of the two groups during both the
training phase and a subsequent testing phase. The primary theoretical
interest is often to assess the influence of training variability on
generalization to novel testing items or conditions. However, the test
may also include some or all of the items that were used during the
training stage, allowing for an assessment of whether the variability
manipulation influenced the learning of the trained items themselves, or
to make it easy to measure how much performance degrades as a function
of how far away testing items are from the training items.

The influence of training variation has received a large amount of
attention in the domain of sensorimotor skill learning. Much of this
research has been influenced by the work of
\textcite{schmidtSchemaTheoryDiscrete1975}, who proposed a schema-based
account of motor learning as an attempt to address the longstanding
problem of how novel movements are produced. Schema theory presumes that
learners possess general motor programs for a class of movements
(e.g.~an underhand throw). When called up for use motor programs are
parameterized by schema rules which determine how the motor program is
parameterized or scaled to the particular demands of the current task.
Schema theory predicts that variable training facilitates the formation
of more robust schemas, which will result in improved generalization or
transfer. Experiments that test this hypothesis are often designed to
compare the transfer performance of a constant-trained group against
that of a varied-trained group. Both groups train on the same task, but
the varied group practices with multiple instances along some
task-relevant dimension that remains invariant for the constant group.
For example, studies using a projectile throwing task might assign
participants to either constant training that practicing throwing from a
single location, and a varied group that throws from multiple locations.
Following training, both groups are then tested from novel throwing
locations
\autocite{pigottMotorSchemaStructure1984,willeyLimitedGeneralizationVaried2018,pachecoLearningSpecificIndividual2018,wulfEffectTypePractice1991}.

One of the earliest, and still often cited investigations of Schmidt's
benefits of variability hypothesis was the work of
\textcite{kerrSpecificVariedPractice1978}. Two groups of children, aged
8 and 12, were assigned to either constant or varied training of a bean
bag throwing task. The constant group practiced throwing a bean-bag at a
small target placed 3 feet in front of them, and the varied group
practiced throwing from a distance of both 2 feet and 4 feet.
Participants were blindfolded and unable to see the target while making
each throw but would receive feedback by looking at where the beanbag
had landed in between each training trial. 12 weeks later, all of the
children were given a final test from a distance of 3 feet which was
novel for the varied participants and repeated for the constant
participants. Participants were also blindfolded for testing and did not
receive trial by trial feedback in this stage. In both age groups,
participants performed significantly better in the varied condition than
the constant condition, though the effect was larger for the younger,
8-year-old children. This result offers a particularly compelling
example of the merits of varied practice, given that the varied group
was able to outperform the constant group even from the home turf
location where one may have expected the constatn group to have the
strongest advantage. A similar pattern of results was observed in
another study wherein varied participants trained with tennis, squash,
badminton, and short-tennis rackets were compared against constant
subjects trained with only a tennis racket
\autocite{greenPracticeVariabilityTransfer1995a}. One of the testing
conditions had subjects repeat the use of the tennis racket, which had
been used on all 128 training trials for the constant group, and only 32
training trials for the varied group. Nevertheless, the varied group
outperformed the constant group when using the tennis racket at testing,
and also performed better in conditions with several novel racket
lengths. However, as is the case with many of the patterns commonly
observed in the ``benefits of variability'' literature, the pattern
wherein the varied group outperfroms the constant group even from the
constants group's home turf has not been consistently replicated. One
recent study attempted a near replication of the Kerr \& Booth study
\autocite{willeyLongtermMotorLearning2018}, having subjects throw
beanbags at a target, with the varied group training from positions (5
and 9 feet) on either side of the constant group (7 feet). This study
did not find a varied advantage from the constant training position,
though the varied group did perform better at distances novel to both
groups. However, this study diverged from the original in that the
participants were adults; and the amount of training was much greater
(20 sessions with 60 practice trials each, spread out over 5-7 weeks).

Pitting varied against constant practice against each other on the home
turf of the constant group provides a compelling argument for the
benefits of varied training, as well as an interesting challenge for
theoretical accounts that posit generalization to occur as some function
of distance. However, despite its appeal this particular contrast is
relatively uncommon in the literature. It is unclear whether this may be
cause for concern over publication bias, or just researchers feeling the
design is too risky. A far more common design is to have separate
constant groups that each train exclusively from each of the conditions
that the varied group encounters
\autocite{catalanoDistantTransferCoincident1984a,chuaPracticeVariabilityPromotes2019,newellVariabilityPracticeTransfer1976,moxleySchemaVariabilityPractice1979,mccrackenTestSchemaTheory1977},
or for a single constant group to train from just one of the conditions
experienced by the varied participants
\autocite{pigottMotorSchemaStructure1984,rollerVariablePracticeLenses2001,wrisbergTrainingProductionNovel1984,wrisbergDevelopingCoincidentTiming1983}.
A less common contrast places the constant group training in a region of
the task space outside of the range of examples experienced by the
varied group, but distinct from the transfer condition
\autocite{wrisbergVariabilityPracticeHypothesis1987,wulfVariabilityPracticeImplicit1997}.
Of particular relevance to the current work is the early study of
\textcite{catalanoDistantTransferCoincident1984a}, as theirs was one of
the earliest studies to investigate the influence of varied vs.~constant
training on multiple testing locations of graded distance from the
training condition. Participants were trained on coincident timing task,
in which subjects observe a series of lightbulbs turning on sequentially
at a consistent rate and attempt to time a button response with the
onset of the final bulb. The constant groups trained with a single
velocity of either 5,7,9, or 11 mph, while the varied group trained from
all 4 of these velocities. Participants were then assigned to one of
four possible generalization conditions, all of which fell outside of
the range of the varied training conditions -- 1, 3, 13 or 15 mph. As is
often the case, the varied group performed worse during the training
phase. In the testing phase, the general pattern was for all
participants to perform worse as the testing conditions became further
away from the training conditions, but since the drop off in performance
as a function of distance was far less steep for the varied group, the
authors suggested that varied training induced a decremented
generalization gradient, such that the varied participants were less
affected by the change between training and testing conditions.

Benefits of varied training have also been observed in many studies
outside of the sensorimotor domain.
\textcite{goodeSuperiorityVariableRepeated2008} trained participants to
solve anagrams of 40 different words ranging in length from 5 to 11
letters, with an anagram of each word repeated 3 times throughout
training, for a total of 120 training trials. Although subjects in all
conditions were exposed to the same 40 unique words (i.e.~the solution
to an anagram), participants in the varied group saw 3 different
arrangements for each solution-word, such as DOLOF, FOLOD, and OOFLD for
the solution word FLOOD, whereas constant subjects would train on three
repetitions of LDOOF (spread evenly across training). Two different
constant groups were used. Both constant groups trained with three
repetitions of the same word scramble, but for constant group A, the
testing phase consisted of the identical letter arrangement to that seen
during training (e.g.~LDOOF), whereas for constant group B, the testing
phase consisted of a arrangement they had not seen during training, thus
presenting them with a testing situation similar situation to the varied
group. At the testing stage, the varied group outperformed both constant
groups, a particularly impressive result, given that constant group A
had 3 prior exposures to the word arrangement (i.e.~the particular
permutation of letters) which the varied group had not explicitly seen.
However varied subjects in this study did not exhibit the typical
decrement in the training phase typical of other varied manipulations in
the literature, and actually achieved higher levels of anagram solving
accuracy by the end of training than either of the constant groups --
solving 2 more anagrams on average than the constant group. This might
suggest that for tasks of this nature where the learner can simply get
stuck with a particular word scramble, repeated exposure to the
identical scramble might be less helpful towards finding the solution
than being given a different arrangement of the same letters. This
contention is supported by the fact that constant group A, who was
tested on the identical arrangement as they experienced during training,
performed no better at testing than did constant group B, who had
trained on a different arrangement of the same word solution -- further
suggesting that there may not have been a strong identity advantage in
this task.

In the domain of category learning, the constant vs.~varied comparison
is much less suitable. Instead, researchers will typically employ
designs where all training groups encounter numerous stimuli, but one
group experiences a greater number of unique exemplars
\autocite{wahlheimMetacognitiveJudgmentsRepetition2012,nosofskyModelguidedSearchOptimal2019,doyleMetacognitiveMonitoringCategory2016,hoschPriorExperienceVariability2023,brunsteinPreparingNoveltyDiverse2011},
or designs where the number of unique training exemplars is held
constant, but one group trains with items that are more dispersed, or
spread out across the category space
\autocite{posnerGenesisAbstractIdeas1968,homaCategoryBreadthAbstraction1976,huHighvariabilityTrainingDoes2024,bowmanTrainingSetCoherence2020,maddoxStimulusRangeDiscontinuity2011}.

Much of the earlier work in this sub-area trained subjects on artificial
categories, such as dot patterns
\autocite{homaCategoryBreadthAbstraction1976,posnerGenesisAbstractIdeas1968}.
A seminal study by \textcite{posnerGenesisAbstractIdeas1968} trained
participants to categorize artificial dot patterns, manipulating whether
learners were trained with low variability examples clustered close to
the category prototypes (i.e.~low distortion training patterns), or
higher-variability patterns spread further away from the prototype
(i.e.~high-distortion patterns). Participants that received training on
more highly-distorted items showed superior generalization to novel high
distortion patterns in the subsequent testing phase. It should be noted
that unlike the sensorimotor studies discussed earlier, the
\textcite{posnerGenesisAbstractIdeas1968} study did not present
low-varied and high-varied participants with an equal number of training
rathers, but instead had participants remain in the training stage of
the experiment until they reached a criterion level of performance. This
train-until-criterion procedure led to the high-variability condition
participants tending to complete a larger number of training trials
before switching to the testing stage. More recent work
\autocite{huHighvariabilityTrainingDoes2024}, also used dot pattern
categories, but matched the number of training trials across conditions.
Under this procedure, higher-variability participants tended to reach
lower levels of performance by the end of the training stage. The
results in the testing phase were the opposite of
\textcite{posnerGenesisAbstractIdeas1968}, with the low-variability
training group showing superior generalization to novel high-distortion
patterns (as well as generalization to novel patterns of low or medium
distortion levels). However, whether this discrepancy is solely a result
of the different training procedures is unclear, as the studies also
differed in the nature of the prototype patterns used.
\textcite{posnerGenesisAbstractIdeas1968} utilized simpler, recognizable
prototypes (e.g., a triangle, the letter M, the letter F), while
\textcite{huHighvariabilityTrainingDoes2024} employed random prototype
patterns.

Recent studies have also begun utilizing more complex or realistic
sitmuli when assessing the influence of variability on category
learning. \textcite{wahlheimMetacognitiveJudgmentsRepetition2012}
conducted one such study. In a within-participants design, participants
were trained on bird categories with either high repetitions of a few
exemplars, or few repetitions of many exemplars. Across four different
experiments, which were conducted to address an unrelated question on
metacognitive judgements, the researchers consistently found that
participants generalized better to novel species following training with
more unique exemplars (i.e.~higher variability), while high repetition
training produced significantly better performance categorizing the
specific species they had trained on. A variability advantage was also
found in the relatively complex domain of rock categorization
\autocite{nosofskyModelguidedSearchOptimal2019}. For 10 different rock
categories, participants were trained with either many repetitions of 3
unique examples of each category, or few repetitions of 9 unique
examples, with an equal number of total training trials in each group
(the design also included 2 other conditions less amenable to
considering the impact of variation). The high-variability group,
trained with 9 unique examples, showed significantly better
generalization performance than the other conditions.
